---
title: "NYC Taxi Trips - Exploratory Data Analysis"
author: "Anwarat Gurung"
date: today
date-format: "ddd, D MMM YYYY"
format:
    html:
        # table of contents
        toc: true
        toc-location: right
        toc-title: "Report Sections"
        toc-depth: 3
        number-sections: true
        anchor-sections: true
        # figure options
        fig-format: svg
        fig-align: center
        fig-width: 8
        fig-height: 6
        # code block options
        code-fold: true 
        code-copy: true 
        code-summary: "Show Code"
        code-line-numbers: false
        # extras
        theme: journal
        page-layout: article
        df-print: paged
        smooth-scroll: true 
execute:
    warning: false
    message: false
    cache: true
---

This report explores the NYC Green Taxi trip data to identify patterns in trip costs and tipping behaviour across different features such as time, passenger count, and location.

```{r code-setup}
#| include: false

# run setup script to load libraries and file paths
source(here::here("scripts", "00_setup.R"))

# export cleaned and processed taxi trips dataset
TAXI_TRIPS_NYC <- readRDS(file = file.path(CLEAN, "taxi_trips_nyc.rds"))
```

# Key Distributions

View numerical distrbution of key numerical variables using Kernal Density Estimate plots.

::: {.panel-tabset .nav-pills}

## Trip Distance

```{r trip-distance-distribution}

colour <- COLOUR_PALETTE[[4]]

ggplot(
    TAXI_TRIPS_NYC, aes(
        x = trip_distance,
    )
) + 
    
    # KDE plot
    geom_density(
        aes(y = after_stat(count)),       # compare actual counts between vendors (instead of probability density)
        adjust = 1.6,       # higher value smoothens fluctuations
        linewidth = 0.7,      # KDE line thickness
        colour = colour,
        fill = colour,
        alpha = 0.2
    ) +
    
    # set x and y axis tick intervals
    scale_x_continuous(n.breaks = 10) +
    scale_y_continuous(n.breaks = 5) +
    xlim(c(NA, 5)) +

    # set text for: title, subtitle, caption
    labs(
        title = "<b>Distribution of Taxi Trip Distances in New York City</b>",
        subtitle = "Count distribution of <b>taxi trip distances</b>, limited to <b>0 - 5 miles</b>, NYC (2021)",
        caption = FIG_CAPTION,
        x = "Trip Distance (miles)", 
        y = NULL    
    ) + 
    
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE 
    ) +
  
    # finetune theme elements
    theme(
        # title, subtitle, caption
        plot.title = element_markdown(          # markdown elements to add HTML/CSS styling/colouring
            size = rel(1.5),                    # relative font size to base_size (from base theme)
            margin = margin(b = 12)             # bottom vertical spacing
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",               # grey
            size = rel(1),
            margin = margin(b = 24),
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1                           # caption right-aligned (0 = left-aligned)
        ),
        plot.title.position = "plot",           # align title to the left of the figure
        # axes, grid, legend, margin
        axis.title.x = element_text(margin = margin(t = 9)),
        panel.grid.major.y = element_blank(),   
        panel.grid.minor = element_blank(),     # remove all minor grid lines
        axis.text.x = element_text(             # x axis ticker labels
            vjust = 0.5,   # distance to x axis
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",               # remove all legends
        plot.margin = margin(15, 15, 15, 15),   # set consistent outer border space
    )
```

## Total Fare Amount

```{r total-fare-amount-distribution}

fare_amount_colour <- COLOUR_PALETTE[[6]]
total_amount_colour <- COLOUR_PALETTE[[2]]

ggplot(
    TAXI_TRIPS_NYC, aes(
        x = fare_amount,
    )
) + 
    
    # KDE plot
    geom_density(
        aes(y = after_stat(count)),     # compare actual counts between vendors (instead of probability density)
        adjust = 2,                   # higher value smoothens fluctuations
        linewidth = 0.7,                  # KDE line thickness
        colour = fare_amount_colour
    ) +

    # KDE plot 
    geom_density(
        aes(
            x = total_amount,
            y = after_stat(count)
        ),       
        adjust = 2,       
        linewidth = 0.7,      
        colour = total_amount_colour,
        alpha = 0.4
    ) +
    
    # set x and y axis tick intervals
    scale_x_continuous(
        n.breaks = 10,
        label = ~ scales::dollar(.)
    ) +
    scale_y_continuous(n.breaks = 5) +
    xlim(c(NA, 50)) +

    # set text for: title, subtitle, caption
    labs(
        title = str_c(
            "Distribution of Taxi ",
            "<span style='color:", fare_amount_colour, ";'><b>Fare Amount</b></span>",
            " and ",
            "<span style='color:", total_amount_colour, ";'><b>Total Amount</b></span>",
            " in New York City"
        ),
        subtitle = "Count distribution across taxi trips in New York City (2021)",
        caption = FIG_CAPTION,
        x = "Fare Amount (USD)", 
        y = NULL    
    ) +
    
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE 
    ) +
  
    # finetune theme elements
    theme(
        # title, subtitle, caption
        plot.title = element_markdown(          # markdown elements to add HTML/CSS styling/colouring
            size = rel(1.5),                    # relative font size to base_size (from base theme)
            margin = margin(b = 12)             # bottom vertical spacing
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",               # grey
            size = rel(1),
            margin = margin(b = 24),
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1                           # caption right-aligned (0 = left-aligned)
        ),
        plot.title.position = "plot",           # align title to the left of the figure
        # axes, grid, legend, margin
        axis.title.x = element_text(margin = margin(t = 9)),
        panel.grid.major.y = element_blank(),   
        panel.grid.minor = element_blank(),     # remove all minor grid lines
        axis.text.x = element_text(             # x axis ticker labels
            vjust = 0.5,   # distance to x axis
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",               # remove all legends
        plot.margin = margin(15, 15, 15, 15),   # set consistent outer border space
    )
```

## Duration

```{r trip-duration-distribution}

colour <- COLOUR_PALETTE[[6]]

ggplot(
    TAXI_TRIPS_NYC, 
    aes(
        x = duration_mins,
    )
) + 
    # KDE plot 
    geom_density(
        aes(y = after_stat(count)),     # compare actual counts between vendors (instead of probability density)
        adjust = 2,                     # higher value smoothens fluctuations
        linewidth = 0.7,                # KDE line thickness
        colour = colour,
        fill = colour,
        alpha = 0.2
    ) +
    
    # set x and y axis tick intervals
    scale_x_continuous(breaks = seq(0, 60, by = 10)) +
    scale_y_continuous(n.breaks = 4) +
    coord_cartesian(
        xlim = c(NA, 60)  # crop x axis limits (does not remove data points)
    ) +

    # set text for: title, subtitle, caption
    labs(
        title = "<b>Distribution of Taxi Trip Durations in New York City</b>",
        subtitle = str_c(
            "Count distribution of <b>trip durations</b> ",
            "across taxi trips in New York City (2021)"
        ),
        caption = FIG_CAPTION,
        x = "Trip Duration (minutes)", 
        y = NULL    
    ) +
    
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE 
    ) +
  
    # finetune theme elements
    theme(
        # title, subtitle, caption
        plot.title = element_markdown(          # markdown elements to add HTML/CSS styling/colouring
            size = rel(1.5),                    # relative font size to base_size (from base theme)
            margin = margin(b = 12)             # bottom vertical spacing
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",               # grey
            size = rel(1),
            margin = margin(b = 24),
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1                           # caption right-aligned (0 = left-aligned)
        ),
        plot.title.position = "plot",           # align title to the left of the figure
        # axes, grid, legend, margin
        axis.title.x = element_text(margin = margin(t = 9)),
        panel.grid.major.y = element_blank(),   
        panel.grid.minor = element_blank(),     # remove all minor grid lines
        axis.text.x = element_text(             # x axis ticker labels
            vjust = 0.5,   # distance to x axis
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",               # remove all legends
        plot.margin = margin(15, 15, 15, 15),   # set consistent outer border space
    )
```

## Overall Hourly Trip Activity

```{r overall-hourly-pickup-kde}
#| code-line-numbers: true

colour <- COLOUR_PALETTE[[8]]

ggplot(
    TAXI_TRIPS_NYC, aes(
        x = pickup_hour,
    )
) + 
    
    # KDE plot with curved vendor text labels (see geomtextpath library)
    geom_density(
        aes(y = after_stat(count)),       # compare actual counts between vendors (instead of probability density)
        adjust = 1.6,       # higher value smoothens fluctuations
        linewidth = 0.7,      # KDE line thickness
        colour = colour,
        fill = colour,
        alpha = 0.2
    ) +
    
    # set x and y axis tick intervals
    scale_x_continuous(
        breaks = seq(0, 23, by = 3),
        labels = ~ sprintf("%02d:00", .)    # format pickup hour 
    ) +
    scale_y_continuous(n.breaks = 6) +
    coord_cartesian(
        xlim = c(NA, 22),   # crop x axis limits (does not remove data points)
        clip = "off"        # prevent clipping of any labels
    ) +
  
    # set text for: title, subtitle, caption
    labs(
        title = "<b>Hourly Patterns in Taxi Demand Across New York City</b>",
        subtitle = "Hourly count distribution of <b>taxi trips</b>, NYC (2021)",
        caption = FIG_CAPTION,
        x = "Pickup Hour", 
        y = NULL    
    ) +
    
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE 
    ) +
  
    # finetune theme elements
    theme(
        # title, subtitle, caption
        plot.title = element_markdown(          # markdown elements to add HTML/CSS styling/colouring
            size = rel(1.5),                    # relative font size to base_size (from base theme)
            margin = margin(b = 12)             # bottom vertical spacing
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",               # grey
            size = rel(1),
            margin = margin(b = 24),
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1                           # caption right-aligned (0 = left-aligned)
        ),
        plot.title.position = "plot",           # align title to the left of the figure
        # axes, grid, legend, margin
        axis.title.x = element_text(margin = margin(t = 9)),
        panel.grid.major.y = element_blank(),   
        panel.grid.minor = element_blank(),     # remove all minor grid lines
        axis.text.x = element_text(             # x axis ticker labels
            vjust = 0.5,   # distance to x axis
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",               # remove all legends
        plot.margin = margin(15, 15, 15, 15),   # set consistent outer border space
    )
```

:::

# Summary of Variables

## Missing Values

```{r}

```

::: panel-tabset

## Numerical Variables

```{r}

```

## Character Variables

```{r}

```

:::