---
title: "NYC Taxi Trips - Data Analysis in R"
author: "Anwarat Gurung"
date: today
date-format: "ddd, D MMM YYYY"
format:
    html:
        # table of contents
        toc: true
        toc-location: right
        toc-title: "Report Sections"
        toc-depth: 3
        number-sections: true
        anchor-sections: true
        # figure options
        fig-format: svg
        fig-align: center
        fig-width: 8
        fig-height: 6
        # code block options
        code-fold: true 
        code-copy: true 
        code-summary: "Show Code"
        code-line-numbers: false
        # extras
        theme: journal
        page-layout: article
        df-print: paged
        smooth-scroll: true 
execute:
    warning: false
    message: false
---

This report presents analyses and visualisations of 2021 taxi trip data from the processed dataset available at `data/clean/taxi_trips_nyc.csv`. The outputs focus on trip activity and patterns, including taxi vendor performance, revenue trends, tipping behaviour, and hourly peak travel periods.

```{r code-setup}
#| include: false

# run setup script to load libraries and file paths
source(here::here("scripts", "00_setup.R"))

# export cleaned and processed taxi trips dataset
TAXI_TRIPS_NYC <- readRDS(file = file.path(CLEAN, "taxi_trips_nyc.rds"))
```


# Revenue by Taxi Vendor

**Which taxi vendor accumulated the most revenue in 2021?**

While revenue per mile is slightly higher for *Green Taxi*, *Uber* recorded significantly greater trip volume and total revenue in 2021, along with higher revenue per trip. 

The full table of metrics is shown below the figure. 

```{r revenue-per-mile-bar-chart}

# aggreagate and summarise data
metrics_by_vendor <- TAXI_TRIPS_NYC %>% 
    group_by(vendorid) %>% 
    summarise(
        # aggregate key metrics
        total_revenue = sum(total_amount),
        total_trips = n(),
        total_distance = sum(trip_distance),
        total_mins = sum(duration_mins)
    ) %>% 
    mutate(
        revenue_per_trip = total_revenue / total_trips,
        revenue_per_mile = total_revenue / total_distance,
        revenue_per_min = total_revenue / total_mins
    ) %>% 
    arrange(desc(revenue_per_mile))

# bar plot

# extract vendor with highest revenue_per_mile
top_vendor <- metrics_by_vendor %>% 
  slice_max(order_by = revenue_per_mile, n = 1) %>% 
  pull(vendorid) %>% 
  as.character()

# assign fill colours for bars
colours <- paletteer_d("futurevisions::hd")
highlight_colour <- COLOUR_GREEN

metrics_by_vendor <- metrics_by_vendor %>% 
  mutate(bar_colour = if_else(vendorid == top_vendor, highlight_colour, "lightgrey"))

# plot: bar chart
ggplot(
    metrics_by_vendor, 
    aes(
        x = fct_reorder(
            vendorid, 
            revenue_per_mile,
            .desc = TRUE
        ),
        y = revenue_per_mile
    )
) +
    # bar chart + highlight colouring
    geom_bar(
        stat = "identity",
        width = .65,
        aes(fill = vendorid == top_vendor)
    ) +
    scale_fill_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "lightgrey"
        )
    ) + 
    
    # text labels for revenue_per_mile
    geom_text(
        aes(
            label = scales::dollar(revenue_per_mile),
            colour = vendorid == top_vendor,
            fontface = ifelse(vendorid == top_vendor, "bold", "plain")
        ),
        size = 5,
        vjust = -1
    ) + 
    scale_colour_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "black"
        )
    ) +
  
    # prevent clipping of any labels
    coord_cartesian(clip = "off") +
  
    # set text for: title, subtitle, caption
    labs(
        title = str_c(
            "<b>Highest Revenue Per Mile: ",
            "<span style='color:", highlight_colour, 
            "; font-weight:bold;'>", top_vendor, "</span>"
        ),
        subtitle = "Average <b>revenue per mile</b> generated by each NYC taxi vendor (2021)",
        caption = FIG_CAPTION,
        x = NULL, y = NULL
    ) +
        
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE
    ) +
  
    # finetune theme elements
    theme(
        # title, subtitle, caption
        plot.title = element_markdown(
            size = rel(1.5),
            margin = margin(b = 12)
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",
            size = rel(1),
            margin = margin(b = 30),
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1
        ),
        plot.title.position = "plot",
        
        # axes, grid, legend, margin
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(
            vjust = 3.5,
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",
        plot.margin = margin(15, 15, 15, 15)
    )
```

A summary of metrics including the total number of rides across taxi vendors, excluding all **voided trips**.

```{r}

# complete Metrics Report

metrics_long <- 
    
    TAXI_TRIPS_NYC %>% 
    
    filter_out(paymenttype == "Voided Trip") %>%    # remove all voided trips
    
    group_by(vendorid) %>% 
    summarise(
        # aggregate key metrics
        total_trips = n(),
        total_distance = sum(trip_distance),
        total_mins = sum(duration_mins),
        total_revenue = sum(total_amount)
    ) %>% 
    mutate(
        revenue_per_trip = total_revenue / total_trips,
        revenue_per_mile = total_revenue / total_distance,
        revenue_per_min = total_revenue / total_mins
    ) %>% 
    arrange(desc(revenue_per_mile)) %>% 

    # transpose all metrics into one column
    pivot_longer(
        cols = -vendorid,
        names_to = "Metric",
        values_to = "Value"
    ) %>% 
    
    # transpose each vendorid into separate columns
    pivot_wider(
        names_from = vendorid,
        values_from = Value
    ) %>% 
    
    # format for presenting
    mutate(
        Metric = Metric %>% 
            str_replace_all("_", " ") %>% 
            str_to_title(),
        across(where(is.numeric), ~ round(., 3)),
    )

metrics_long %>% 
    slice(1:4) %>% 
    mutate(
        across(where(is.numeric), ~ scales::comma(.)), 
    ) %>% 
    kable()
```

Revenue-based metrics:

```{r}
#| echo: false 

metrics_long %>% 
    slice(5:7) %>% 
    mutate(
        across(where(is.numeric), ~ scales::dollar(.)), 
    ) %>% 
    kable()
```

## Daily Revenue Trends

**How has revenue changed over 2021 for both vendors?**

Revenue for both vendors remained broadly stable throughout 2021, with short-term fluctuations, while `Uber` showed a slight upward trend.

::: panel-tabset

### Plot

```{r daily-rev-series}
#| code-line-numbers: true

# summarise daily revenues for Green Taxi and Uber only
daily_revenues <- TAXI_TRIPS_NYC %>% 
    filter(vendorid %in% c("Green Taxi", "Uber")) %>% 
    group_by(vendorid, pickup_date) %>%
    summarise(
        daily_rev = sum(total_amount), 
        .groups = "drop"
    )     

ggplot(
    data = daily_revenues,
    mapping = aes(
        x = pickup_date,
        y = daily_rev,
        colour = vendorid       # group by vendorid for separate lines
    )
) +
    # line plot (series)
    geom_line(
        # alpha = 0.3,
        linewidth = 0.6
    ) +
    
    # prevent clipping of any labels
    coord_cartesian(clip = "off") +        

# hardcoded date filtering
    scale_x_date(
        limits = as.Date(c("2021-07-01", "2021-07-31"))
    ) +
    
    geom_smooth(
        method = "lm",
        se = FALSE,
        linetype = "dashed",
        alpha = 0.4,
        linewidth = 0.6
    ) + 
    
    scale_y_continuous(
        n.breaks = 10,
        labels = scales::label_number(
            prefix = "$",
            scale = 1e-3,
            suffix = "k",
            accuracy = 1
        )
    ) +
    
    # manually set colour for each taxi vendor
    scale_colour_manual(
        values = c(
            "Green Taxi" = COLOUR_GREEN,
            "Uber" = COLOUR_UBER
        )
    ) +

    # set text for: title, subtitle, caption
    labs(
        title = str_c(
            "<b>Daily Taxi Revenue in NYC: ",
            "<span style='color:", COLOUR_UBER, "; font-weight:bold;'>Uber</span> vs ",
            "<span style='color:", COLOUR_GREEN, "; font-weight:bold;'>Green Taxi</span></b>"
        ),
        subtitle = "Daily revenue from taxi trips in NYC by vendor",
        caption = FIG_CAPTION,
        x = NULL, y = NULL
    ) +
    
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE 
    ) +
  
    # finetune theme elements
    theme(
        plot.title = element_markdown(          # markdown elements to add HTML/CSS styling/colouring
            size = rel(1.5),                    # relative font size to base_size (from base theme)
            margin = margin(b = 12)             # bottom vertical spacing
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e", # grey
            size = rel(1),
            margin = margin(b = 24),
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1,                          # caption right-aligned (0 = left-aligned)
            margin = margin(t = 12),
        ),
        plot.title.position = "plot",           # align title to the left of the figure
        axis.title.x = element_text(margin = margin(t = 8)),
        # panel.grid.major.x = element_blank(),   
        panel.grid.minor = element_blank(),     # remove all minor grid lines
        axis.text.x = element_text(             # x axis ticker labels
            vjust = 1,   # distance to x axis
            colour = "black",
            size = rel(1.2)
        ),
        axis.text.y = element_text(             # x axis ticker labels
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",               # remove all legends
        plot.margin = margin(15, 15, 15, 15),   # set consistent outer border space
    )
```

### Data

```{r}
#| code-fold: false

daily_revenues
```

:::

```{r}
#| include: false

# save figure output to file
ggsave(
    file.path(FIGURES, "revenue_trends_by_vendor.svg"),
    width  = opts_current$get("fig.width"),
    height = opts_current$get("fig.height")
)
```

## Green Taxi Revenue by Trip Type

**Which trip type has brought the most revenue to Green Taxi?**

*Inner City* trips generated the vast majority of Green Taxi revenue, with *Outer City* trips contributing only a negligible share.

::: panel-tabset

### Plot
```{r rev-by-trip-type-bar}
#| code-line-numbers: true

# aggregate and summarise data
rev_by_trips <- TAXI_TRIPS_NYC %>% 
    filter(vendorid == "Green Taxi" & !is.na(trip_type)) %>% 
    group_by(trip_type) %>% 
    summarise(
        total_revenue = sum(total_amount),
        # optional: extra insights
        total_trips = n(),
        total_mins = sum(duration_mins)
    ) %>% 
    mutate(revenue_per_min = total_revenue / total_mins)

# extract the trip_type with the highest revenue
top_trip_type <- rev_by_trips %>% 
    slice_max(order_by = total_revenue, n = 1) %>% 
    pull(trip_type) %>%      
    as.character()   

# assign fill colours for bars
highlight_colour <- "#751C6DFF"               
rev_by_trips <- mutate(rev_by_trips, bar_colour = if_else(trip_type == top_trip_type, highlight_colour, "lightgrey"))

# plot: bar chart
ggplot(
    rev_by_trips, 
    aes(
        x = fct_reorder(
            trip_type, 
            total_revenue,    
            .desc = TRUE
        ),
        y = total_revenue
    )
) +
    # bar chart + highlight colouring
    geom_bar(
        stat = "identity",
        width = 0.5,
        aes(fill = trip_type == top_trip_type),  
    ) +
    scale_fill_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "lightgrey"
        )
    ) + 
    
    # text labels (of revenue values)
    geom_text(
        aes(
            label = scales::label_dollar(accuracy = 1)(total_revenue),
            colour = trip_type == top_trip_type,
            fontface = ifelse(trip_type == top_trip_type, "bold", "plain")
        ),
        size = 5,
        vjust = -1
    ) + 
    scale_colour_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "black"
        )
    ) +
  
    # set text for: title, subtitle, caption
    labs(
        title = str_c(
            "<b>The Majority of Green Taxi Revenue is Generated by ",
            "<span style='color:", highlight_colour, 
            "; font-weight:bold;'>", top_trip_type, 
            "</span> Trips"
        ),
        subtitle = "Total revenue of <b>Green Taxi</b> trips by trip type, NYC (2021)",
        caption = FIG_CAPTION,
        x = NULL, y = NULL    
    ) +
        
    # prevent clipping of any labels
    coord_cartesian(clip = "off") +
    
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE
    ) +
    
    # finetune theme elements
    theme(
        plot.title = element_markdown(
            size = rel(1.5),
            margin = margin(b = 12),
            lineheight = 1.4
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",
            size = rel(1),
            margin = margin(b = 30)
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1
        ),
        plot.title.position = "plot",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(
            vjust = 3.5,
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",
        plot.margin = margin(15, 15, 15, 15)
    )
```

### Data

```{r}
#| code-fold: false

rev_by_trips %>% select(-bar_colour) %>% kable()
```

```{r}
#| include: false
ggsave(
    file.path(FIGURES, "gt_revenue_by_trip_type.svg"),
    width  = opts_current$get("fig.width"),
    height = opts_current$get("fig.height")
)
```

:::

# Average Total Green Taxi Spend by Weekday

**What is the average amount spent on Green Taxi trips across the week, and are any weekdays significantly higher than others?**

On average, total amounts were slightly higher on Sunday and Thursday, with the lowest on Saturday, but differences across the week were very minimal.

::: panel-tabset

## Plot 

```{r avg-spend-by-weekday}

# aggregate and summarise data
avg_spend_by_weekday <-  TAXI_TRIPS_NYC %>% 
    filter(vendorid == "Green Taxi") %>% 
    group_by(pickup_weekday) %>% 
    summarise(avg_amount = mean(total_amount))

# identify weekday with highest average spend
top_weekday <- avg_spend_by_weekday %>% 
    slice_max(avg_amount, n = 1) %>% 
    pull(pickup_weekday) %>% 
    as.character()

# assign fill colours for bars
colours <- paletteer_d("nationalparkcolors::Badlands")      # pick a colour palette
highlight_colour <- colours[[4]]                            # pick a highlight colour from palette
avg_spend_by_weekday <- avg_spend_by_weekday %>% 
  mutate(bar_colour = if_else(pickup_weekday == top_weekday, highlight_colour, "lightgrey"))

# plot: bar chart
ggplot(
    avg_spend_by_weekday, 
    aes(
        # horizontal bar chart layout
        y = fct_rev(pickup_weekday),    # order weekday from Mon - Sun
        x = avg_amount
    )
) +

    # bar chart + highlight colouring
    geom_bar(
        stat = "identity",       # use directly aggregated values
        width = 0.8,
        # map fill to a bool condition to highlight the top borough
        aes(fill = pickup_weekday == top_weekday),  
    ) +   
    scale_fill_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "lightgrey"
        )
    ) +
    
    # add text labels to show values
    geom_text(
        aes(label = scales::label_currency(prefix = "$")(avg_amount),
            # map fill to a bool condition to highlight the top value
            colour = pickup_weekday == top_weekday,
            # repeat to make top borough bold font
            fontface = ifelse(pickup_weekday == top_weekday, "bold", "plain")
        ),
        size = 4.5,
        hjust = 1.2
    ) + 
    # add weekday text labels
    geom_text(  
        aes(
            label = pickup_weekday,
            colour = pickup_weekday == top_weekday,
            fontface = ifelse(pickup_weekday == top_weekday, "bold", "plain")
        ),
        x = 0.4,     
        size = 4.5,    
        hjust = 0
    ) +
    scale_colour_manual(
        values = c(
            "TRUE" = "white",
            "FALSE" = "black"
        )
    ) +
  
    # set text for: title, subtitle, caption
    labs(
        title = str_c(
            "<b>Weekday with the Highest Average Trip Cost: ",
            "<span style='color:", highlight_colour, 
            "; font-weight:bold;'>", top_weekday, "</span>"
        ),
        subtitle = "Average total amount paid per <b>Green Taxi</b> trip by weekday, NYC (2021)",
        caption = "Source: NYC Taxi & Limousine Commission (TLC) Trip Record Data",
        x = NULL, y = NULL    # no x/y axis label for bar charts
    ) +
    
    # crop x and y limits (does not remove any data points)
    coord_cartesian(
        # xlim = c(10, NA),       # used to re-scale proportions
        clip = "off"            # prevent clipping of any labels
    ) +     
        
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE   
    ) +
  
    # finetune theme elements
    theme(

        # title, subtitle, caption
        plot.title = element_markdown(          # markdown elements to add HTML/CSS styling/colouring
            size = rel(1.5),                    # relative font size to base_size (from base theme)
            margin = margin(b = 16)             # bottom vertical spacing
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",               # grey
            size = rel(1),
            margin = margin(b = 12),
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1                           # caption right-aligned (0 = left-aligned)
        ),
        plot.title.position = "plot",           # align title to the left of the figure

        # axes, grid, legend, margin
        axis.text.y = element_blank(),          # remove original weekday ticker labels
        axis.text.x = element_blank(),          # remove x axis ticker
        panel.grid = element_blank(),           # remove all grid lines
        legend.position = "none",                # remove all legends
        plot.margin = margin(15, 15, 15, 15),   # set consistent outer border space
    )
```

## Data 

```{r}
#| code-fold: false 

avg_spend_by_weekday %>% select(-bar_colour) %>% kable()
```

```{r}
#| include: false

# save figure output to file
ggsave(
    file.path(FIGURES, "avg_spend_by_weekday.svg"),
    width  = opts_current$get("fig.width"),
    height = opts_current$get("fig.height")
)
```

:::

# Pickup Borough Distribution

**Which pickup borough location is the most popular for Green Taxi?**

*Manhattan* is the dominant pickup borough for Green Taxi, followed by *Brooklyn*, *Queens* and *Bronx*.

::: panel-tabset

## Green Taxi Pickups

```{r trips-by-borough-bar}
#| code-line-numbers: true

# aggregate and summarise data
trips_by_borough <- TAXI_TRIPS_NYC %>% 
    filter(vendorid == "Green Taxi" & !is.na(pickup_borough)) %>% 
    group_by(pickup_borough) %>% 
    summarise(total_trips = n())   # count no. of trips per borough   

# extract the most popular pickup_borough by total_trips
top_borough <- trips_by_borough %>% 
  slice_max(order_by = total_trips, n = 1) %>% 
  pull(pickup_borough) %>%      # return single column
  as.character()                # convert to text character

# assign fill colours for bars
colours <- paletteer_d("futurevisions::hd")     # pick a colour palette
highlight_colour <- colours[[1]]                # pick a highlight colour from palette
trips_by_borough <- trips_by_borough %>% 
  mutate(bar_colour = if_else(pickup_borough == top_borough, highlight_colour, "lightgrey"))

# plot: bar chart
ggplot(
    trips_by_borough, 
    aes(
        x = fct_reorder(
            pickup_borough, 
            total_trips,    # order by descending trip counts
            .desc = TRUE
        ),
        y = total_trips
    )
) +
    # bar chart + highlight colouring
    geom_bar(
        stat = "identity",
        width = 0.8,
        # map fill to a bool condition to highlight the top borough
        aes(fill = pickup_borough == top_borough),  
    ) +
    scale_fill_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "lightgrey"
        )
    ) + 
    
    # text labels (of values) for total_trips
    geom_text(
        aes(
            label = scales::comma(total_trips),
            # map fill to a bool condition to highlight the top value
            colour = pickup_borough == top_borough,
            # repeat to make top borough bold font
            fontface = ifelse(pickup_borough == top_borough, "bold", "plain")
        ),
        size = 5,
        vjust = -1
    ) + 
    scale_colour_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "black"
        )
    ) +
  
    # prevent clipping of any labels
    coord_cartesian(clip = "off") +
  
    # set text for: title, subtitle, caption
    labs(
        title = str_c(
            "<b>Top Pickup Borough For Green Taxis: ",
            "<span style='color:", highlight_colour, 
            "; font-weight:bold;'>", top_borough, "</span>"
        ),
        subtitle = "Total number of <b>Green Taxi</b> trips by pickup borough, NYC (2021)",
        caption = FIG_CAPTION,
        x = NULL, y = NULL    # no x/y axis label for bar charts
    ) +
        
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE
    ) +
  
    # finetune theme elements
    theme(
        # title, subtitle, caption
        plot.title = element_markdown(          # markdown elements to add HTML/CSS styling/colouring
            size = rel(1.5),                    # relative font size to base_size (from base theme)
            margin = margin(b = 12)             # bottom vertical spacing
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",               # grey
            size = rel(1),
            margin = margin(b = 24),
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1                           # caption right-aligned (0 = left-aligned)
        ),
        plot.title.position = "plot",           # align title to the left of the figure
        # axes, grid, legend, margin
        panel.grid.major.x = element_blank(),   # remove vertical major lines
        panel.grid.minor = element_blank(),     # remove all minor grid lines
        axis.text.x = element_text(             # x axis ticker labels
            vjust = 3.5,   # distance to x axis
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",               # remove all legends
        plot.margin = margin(15, 15, 15, 15),   # set consistent outer border space
    )
```

```{r}
#| code-fold: false

trips_by_borough %>% select(-bar_colour) %>% kable()
```

```{r}
#| include: false

# save figure output to file
ggsave(
    file.path(FIGURES, "gt_trips_by_borough.svg"),
    width  = opts_current$get("fig.width"),
    height = opts_current$get("fig.height")
)
```

## Competitor Pickups

**Are there any boroughs where Green Taxi is underperforming significantly versus competitors?**

Across all boroughs *Green Taxi* lags significantly behind both *Uber* and *Other* vendors in pickup volume market share.

```{r vendor-trips-by-borough-bar}
#| code-fold: true 
#| code-line-numbers: true

vendor_trips_by_borough <- TAXI_TRIPS_NYC %>% 
    filter(!is.na(pickup_borough)) %>% 
    group_by(vendorid, pickup_borough) %>% 

    # counts no. of trips recorded per borough
    summarise(total_trips = n(), .groups = "drop") %>% 

    group_by(pickup_borough) %>% 
    mutate(borough_total = sum(total_trips)) %>% 

    # optional: reorder vendorid factor based on individual trip counts
    mutate(vendorid = fct_reorder(vendorid, total_trips, .fun = sum, .desc = TRUE))

ggplot(
    vendor_trips_by_borough,
    aes(
        x = fct_reorder(pickup_borough, borough_total, .desc = TRUE),
        y = total_trips,
        fill = vendorid
    )
) +
    geom_bar(
        stat = "identity",
        position = position_dodge(preserve = "single"),
        # alpha = 0.
    ) +
    
    # manually set colour for each taxi vendor
    scale_fill_manual(
        values = c(
            "Green Taxi" = COLOUR_GREEN,
            "Uber" = COLOUR_UBER,
            "Other" = COLOUR_OTHER
        )
    ) +
    
    labs(
    title = str_c(
        "<span style='color:", COLOUR_GREEN, ";'><b>Green Taxi</b></span> ",
        "lags behind competitors ",
        "<span style='color:", COLOUR_UBER, ";'><b>Uber</b></span> ",
        "and ",
        "<span style='color:", COLOUR_OTHER, ";'><b>Other</b></span>"
    ),
    subtitle = "Total recorded taxi pickup trips by vendor across NYC boroughs (2021)",
    caption = FIG_CAPTION,
    x = NULL, y = NULL
    ) +
    
    coord_cartesian(clip = "off") +

    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE
    ) +
    
    theme(
        # title, subtitle, caption
        plot.title = element_markdown(
            size = rel(1.5),
            margin = margin(b = 12)
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",
            size = rel(1),
            margin = margin(b = 16)
        ),
        plot.caption = element_text(
            colour = "#6e6e6e",
            size = rel(0.7),
            hjust = 1
        ),
        plot.title.position = "plot",

        # axes, grid, legend, margin
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(
            vjust = 3.5,
            colour = "black",
            size = rel(1.2)
        ),
        axis.text.y = element_text(
            colour = "black"
        ),
        legend.position = "none",
        plot.margin = margin(15, 15, 15, 15)
    )
```

```{r}
#| code-fold: false

vendor_trips_by_borough
```

```{r}
#| include: false

# save figure output to file
ggsave(
    file.path(FIGURES, "vendor_trips_by_borough.svg"),
    width  = opts_current$get("fig.width"),
    height = opts_current$get("fig.height")
)
```

:::

# Variables Affecting Green Taxi Tips

**Do any features affect tip amount for Green Taxi?**

The following charts explore how different features affect the **tip amount** for Green Taxi trips.

::: {.panel-tabset .nav-pills}

## Pickup Hour

```{r tip-amount-vs-pickup-hour}

avg_tips_by_hour <- TAXI_TRIPS_NYC %>% 
    filter(vendorid == "Green Taxi") %>% 
    select(pickup_hour, tip_amount) %>% 
    group_by(pickup_hour) %>% 
    summarise(avg_tip = mean(tip_amount))

max_avg_tip <- max(avg_tips_by_hour[["avg_tip"]])
min_avg_tip <- min(avg_tips_by_hour[["avg_tip"]])

colour  <- COLOUR_PALETTE[[6]]
highlight_colour <- COLOUR_PALETTE[[1]]

ggplot(
    avg_tips_by_hour,
    aes(
        x = pickup_hour, 
        y = avg_tip
    )
) +
    
# points for each hour
    geom_point(
        colour = colour,
        size = 1.5,
        alpha = 0.5
    ) +
    
    # line connecting points
    geom_line(
        colour = colour,
        linewidth = 0.5,
        alpha = 0.4
    ) +
    
    # polynomial trend line
    geom_smooth(
        method = "lm",
        formula = y ~ poly(x, 4),
        se = FALSE,
        # linetype = "dashed",
        alpha = 0.4,
        linewidth = 0.9,
        colour = colour
    ) +
    
    # highlight max avg_tip
    geom_point(
        data = avg_tips_by_hour %>% filter(avg_tip == max_avg_tip),
        aes(x = pickup_hour, y = avg_tip),
        colour = highlight_colour,
        size = 2
    ) +
    geom_text(
        data = avg_tips_by_hour %>% filter(avg_tip == max_avg_tip),
        aes(
            x = pickup_hour,
            y = avg_tip,
            label = scales::label_currency(prefix = "$")(max_avg_tip)
        ),
        vjust = 2,
        hjust = -0.6,
        colour = highlight_colour,
        fontface = "bold"
    ) +
    
    # highlight min avg_tip
    geom_point(
        data = avg_tips_by_hour %>% filter(avg_tip == min_avg_tip),
        aes(x = pickup_hour, y = avg_tip),
        colour = highlight_colour,
        size = 2
    ) +
    geom_text(
        data = avg_tips_by_hour %>% filter(avg_tip == min_avg_tip),
        aes(
            x = pickup_hour,
            y = avg_tip,
            label = scales::label_currency(prefix = "$")(min_avg_tip)
        ),
        vjust = -1,
        hjust = -0.6,
        colour = highlight_colour,
        fontface = "bold"
    ) +
    
    # x-axis and y-axis formatting
    # set x and y axis tick intervals
    scale_x_continuous(
        breaks = seq(0, 24, by = 4),
        labels = ~ sprintf("%02d:00", .)    # format pickup hour 
    ) +

    scale_y_continuous(
        n.breaks = 8,
        labels = scales::label_number(
            prefix = "$",
            accuracy = 0.01
        )
    ) +
    
    # chart title, subtitle, caption
    labs(
        title = "<b>Average Tip Amount by <b>Pickup Hour</b>",
        subtitle = "Average tip received per hour of the day for Green Taxi trips",
        caption = FIG_CAPTION,
        x = "Pickup Hour", y = NULL
    ) +
    
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE
    ) +
    
    # fine-tune theme elements
    theme(
        plot.title = element_markdown(
            size = rel(1.5),
            margin = margin(b = 12)
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",
            size = rel(1),
            margin = margin(b = 24)
        ),
        plot.caption = element_text(
            colour = "#6e6e6e",
            size = rel(0.7),
            hjust = 1,
            margin = margin(t = 12)
        ),
        plot.title.position = "plot",
        axis.title.x = element_text(margin = margin(t = 8)),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(
            vjust = 1,
            colour = "black",
            size = rel(1.2)
        ),
        axis.text.y = element_text(
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",
        plot.margin = margin(15, 15, 15, 15)
    )
```

```{r}
#| code-fold: false

avg_tips_by_hour
```

## Passenger Count

```{r tip-amount-vs-passenger-count}

# aggregate and summarise data
avg_tips_by_passenger_count <- TAXI_TRIPS_NYC %>% 
    filter(
        vendorid == "Green Taxi",
        !is.na(passenger_count)
    ) %>% 
    select(passenger_count, tip_amount) %>% 
    group_by(passenger_count) %>% 
    summarise(
        avg_tip = mean(tip_amount),
        n_trips = n()
    )

# extract the passenger count with highest avg_tip
top_passenger_count <- avg_tips_by_passenger_count %>% 
  slice_max(order_by = avg_tip, n = 1) %>% 
  pull(passenger_count) %>% 
  as.character()

# assign fill colours for bars
colours <- paletteer_d("futurevisions::hd")     # pick a colour palette
highlight_colour <- colours[[3]]                # pick a highlight colour
avg_tips_by_passenger_count <- avg_tips_by_passenger_count %>% 
  mutate(bar_colour = if_else(passenger_count == top_passenger_count, highlight_colour, "lightgrey"))

# plot: bar chart with styling identical to exemplar
ggplot(
    avg_tips_by_passenger_count,
    aes(
        x = fct_reorder(
            factor(passenger_count),
            avg_tip,
            .desc = TRUE
        ),
        y = avg_tip
    )
) +
    # bar chart + highlight colouring
    geom_bar(
        stat = "identity",
        width = 0.8,
        aes(fill = passenger_count == top_passenger_count)
    ) +
    scale_fill_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "lightgrey"
        )
    ) +
    
    # text labels for avg_tip
    geom_text(
        aes(
            label = scales::label_currency(prefix = "$")(avg_tip),
            colour = passenger_count == top_passenger_count,
            fontface = ifelse(passenger_count == top_passenger_count, "bold", "plain")
        ),
        size = 5,
        vjust = -1
    ) +
    scale_colour_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "black"
        )
    ) +
    
    # prevent clipping of labels
    coord_cartesian(clip = "off") +

    # format y axis as currency
    scale_y_continuous(
        n.breaks = 4,
        labels = scales::label_number(
            prefix = "$",
            accuracy = 0.01
        )
    ) +
    
    # set title, subtitle, caption
    labs(
        title = str_c(
            "<b>Passenger Count with Highest Average Tip: ",
            "<span style='color:", highlight_colour, 
            "; font-weight:bold;'>", top_passenger_count, "</span>"
        ),
        subtitle = "Average tip value by passenger count for <b>Green Taxi</b> trips, NYC (2021)",
        caption = FIG_CAPTION,
        x = "Passenger Count", y = NULL
    ) +
    
    # base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE
    ) +
    
    # fine-tune theme elements
    theme(
        plot.title = element_markdown(
            size = rel(1.5),
            margin = margin(b = 12)
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",
            size = rel(1),
            margin = margin(b = 24)
        ),
        plot.caption = element_text(
            colour = "#6e6e6e",
            size = rel(0.7),
            hjust = 1
        ),
        plot.title.position = "plot",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(
            vjust = 3.5,
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",
        plot.margin = margin(15, 15, 15, 15)
    )
```

```{r}
#| code-fold: false

avg_tips_by_passenger_count %>% select(-bar_colour) %>% kable()
```

## Pickup Borough

```{r tip-amount-vs-pickup-borough}

# aggregate and summarise data
avg_tips_by_borough <- TAXI_TRIPS_NYC %>% 
    filter(
        vendorid == "Green Taxi",
        !is.na(pickup_borough)
    ) %>% 
    select(pickup_borough, tip_amount) %>% 
    group_by(pickup_borough) %>% 
    summarise(
        avg_tip = mean(tip_amount),
        n_trips = n()
    )

# extract the borough with highest avg_tip
top_borough_tip <- avg_tips_by_borough %>% 
  slice_max(order_by = avg_tip, n = 1) %>% 
  pull(pickup_borough) %>% 
  as.character()

# assign fill colours for bars
highlight_colour <- COLOUR_PALETTE[[2]]                
avg_tips_by_borough <- avg_tips_by_borough %>% 
  mutate(bar_colour = if_else(pickup_borough == top_borough_tip, highlight_colour, "lightgrey"))

# plot
ggplot(
    avg_tips_by_borough,
    aes(
        x = fct_reorder(
            pickup_borough,
            avg_tip,
            .desc = TRUE
        ),
        y = avg_tip
    )
) +
    # bar chart with highlight
    geom_bar(
        stat = "identity",
        width = 0.7,
        aes(fill = pickup_borough == top_borough_tip)
    ) +
    scale_fill_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "lightgrey"
        )
    ) +
    
    # text labels: average tip
    geom_text(
        aes(
            label = scales::label_currency(prefix = "$")(avg_tip),
            colour = pickup_borough == top_borough_tip,
            fontface = ifelse(pickup_borough == top_borough_tip, "bold", "plain")
        ),
        size = 5,
        vjust = -1
    ) +
    scale_colour_manual(
        values = c(
            "TRUE" = highlight_colour,
            "FALSE" = "black"
        )
    ) +
    
    # prevent clipping
    coord_cartesian(clip = "off") +
    
    # format y axis as currency
    scale_y_continuous(
        n.breaks = 4,
        labels = scales::label_number(
            prefix = "$",
            accuracy = 0.01
        )
    ) +

    # title, subtitle, caption
    labs(
        title = str_c(
            "<b>Pickup Borough with Highest Average Tip: ",
            "<span style='color:", highlight_colour, 
            "; font-weight:bold;'>", top_borough_tip, "</span>"
        ),
        subtitle = "Average tip amount by pickup borough for <b>Green Taxi</b> trips, NYC (2021)",
        caption = FIG_CAPTION,
        x = NULL, y = NULL
    ) +
    
    # base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE
    ) +
    
    # fine-tune theme elements
    theme(
        plot.title = element_markdown(
            size = rel(1.5),
            margin = margin(b = 12)
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",
            size = rel(1),
            margin = margin(b = 24)
        ),
        plot.caption = element_text(
            colour = "#6e6e6e",
            size = rel(0.7),
            hjust = 1
        ),
        plot.title.position = "plot",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(
            vjust = 4,
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",
        plot.margin = margin(15, 15, 15, 15)
    )

```

```{r}
#| code-fold: false

 avg_tips_by_borough %>% select(-bar_colour) %>% kable()
```

## Trip Distance

```{r tip-amount-vs-trip-distance}

TAXI_TRIPS_NYC %>% 
    filter(
        vendorid == "Green Taxi",
        tip_amount != 0
    ) %>% 
    select(trip_distance, tip_amount) %>% 
    
    # scatter plot
    ggplot(
        aes(
            x = trip_distance,
            y = tip_amount
        )
    ) +
        
    # scatter points
    geom_point(
        colour = COLOUR_PALETTE[7],
        size = 3,
        alpha = 0.4
    ) +

    # filter data points
    scale_x_continuous(limits = c(0, 25)) +
    scale_y_continuous(limits = c(0, 20)) +
    
    # linear fit
    geom_smooth(
        method = "lm",
        se = FALSE,
        size = 0.6,
        colour = COLOUR_PALETTE[[2]],
        alpha = 0.6
    ) +
        
    # labels
    labs(
        title = "Relationship Between  <b>Tip Amount</b> and <b>Trip Distance</b>",
        subtitle = "Tip amount plotted against trip distance for <b>Green Taxi</b> trips, NYC (2021)",
        caption = FIG_CAPTION,
        x = "Trip Distance (miles)",
        y = "Tip Amount ($)"
    ) +
    
    # base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE
    ) +
    
    # fine-tune theme
    theme(
        plot.title = element_markdown(
            size = rel(1.5),
            margin = margin(b = 12)
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",
            size = rel(1),
            margin = margin(b = 24)
        ),
        plot.caption = element_text(
            colour = "#6e6e6e",
            size = rel(0.7),
            hjust = 1,
            margin = margin(t = 12)
        ),
        plot.title.position = "plot",
        # panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(
            colour = "black",
            size = rel(1.2),
            vjust = 2
        ),
        axis.text.y = element_text(
            colour = "black",
            size = rel(1.2),
            hjust = 0
        ),
        plot.margin = margin(15, 15, 15, 15)
    )
```

:::

# Hourly Taxi Trip Activity 

**What are the peak hours for taxi services?**

::: panel-tabset

## Overall Hourly Trip Activity

```{r overall-hourly-pickup-kde}
#| code-line-numbers: true

ggplot(
    TAXI_TRIPS_NYC, aes(
        x = pickup_hour,
    )
) + 
    
    # KDE plot with curved vendor text labels (see geomtextpath library)
    geom_density(
        aes(y = after_stat(count)),       # compare actual counts between vendors (instead of probability density)
        adjust = 1.6,       # higher value smoothens fluctuations
        linewidth = 1,      # KDE line thickness
        colour = COLOUR_PALETTE[[2]]
    ) +

    # extra: annotations for peaks
    annotate(
        geom = "text",
        x = 4.5, y = 4500,
        label = "Morning Peak\nfrom 9-10am",
        hjust = 0.5,
        size = 3.5,
        family = BASE_FAMILY
    ) +
    annotate(
        geom = "curve",
        x = 4.5, y = 4800,
        xend = 9.9, yend = 5650,
        arrow = arrow(
            length = unit(0.25, "cm"),     # arrow head length
            angle = 15,                    # adjust width of arrow head
            ends = "last",
            type = "closed"                # filled arrow head
        ),
        curvature = -0.5,                  # arrow radius
        linewidth = 0.3,                 
        colour = "black",
        size = 4,
        family = BASE_FAMILY
    ) +
    
    # set x and y axis tick intervals
    scale_x_continuous(
        breaks = seq(0, 23, by = 3),
        labels = ~ sprintf("%02d:00", .)    # format pickup hour 
    ) +
    scale_y_continuous(n.breaks = 4) +
    coord_cartesian(
        xlim = c(NA, 22),   # crop x axis limits (does not remove data points)
        clip = "off"        # prevent clipping of any labels
    ) +
  
    # set text for: title, subtitle, caption
    labs(
        title = "<b>Hourly Patterns in Taxi Demand Across New York City</b>",
        subtitle = "Hourly distribution of <b>taxi trips</b>, NYC (2021)",
        caption = FIG_CAPTION,
        x = "Pickup Hour", 
        y = NULL    
    ) +
    
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE 
    ) +
  
    # finetune theme elements
    theme(
        # title, subtitle, caption
        plot.title = element_markdown(          # markdown elements to add HTML/CSS styling/colouring
            size = rel(1.5),                    # relative font size to base_size (from base theme)
            margin = margin(b = 12)             # bottom vertical spacing
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",               # grey
            size = rel(1),
            margin = margin(b = 24),
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1                           # caption right-aligned (0 = left-aligned)
        ),
        plot.title.position = "plot",           # align title to the left of the figure
        # axes, grid, legend, margin
        axis.title.x = element_text(margin = margin(t = 9)),
        panel.grid.major.y = element_blank(),   
        panel.grid.minor = element_blank(),     # remove all minor grid lines
        axis.text.x = element_text(             # x axis ticker labels
            vjust = 0.5,   # distance to x axis
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",               # remove all legends
        plot.margin = margin(15, 15, 15, 15),   # set consistent outer border space
    )
```

```{r}
#| include: false

# save figure output to file
ggsave(
    file.path(FIGURES, "overall_pickup_hour_activity.svg"),
    width  = opts_current$get("fig.width"),
    height = opts_current$get("fig.height")
)
```

## Hourly Trip Activity by Vendor

**Which vendor is most active in the evening?**

*Uber* dominates evening demand and activity from 3pm onwards, with a peak at 6pm, while *Other* vendors dominate the morning peak around 9-10am.

```{r hourly-pickup-by-vendor-kde}
#| code-line-numbers: true

# plot: kernel density estimate (KDE) 
ggplot(
    TAXI_TRIPS_NYC, aes(
        x = pickup_hour,
        colour = vendorid       # group by taxi vendor (line for each one)
    )
) + 
    
    # KDE plot with curved vendor text labels (see geomtextpath library)
    geom_textdensity(
        aes(
            y = after_stat(count),       # compare actual counts between vendors (instead of probability density)
            label = vendorid
        ),     
        adjust = 1.6,           # higher value smoothens fluctuations
        linewidth = 0.7,        # KDE line thickness
        # geomtext path parameters
        hjust = 0.75,
        gap = FALSE,            # don't create gaps within lines
        vjust = -0.25,          # move text labels above respective line
        size = 4,
        fontface = "bold",
        family = BASE_FAMILY    # font family
    ) +
    
    # manually set colour for each taxi vendor
    scale_colour_manual(
        values = c(
            "Green Taxi" = COLOUR_GREEN,
            "Uber" = COLOUR_UBER,
            "Other" = COLOUR_OTHER
        )
    ) +
    
    # set x and y axis tick intervals
    scale_x_continuous(
        breaks = seq(0, 23, by = 3),
        labels = ~ sprintf("%02d:00", .)    # format pickup hour 
    ) +
    scale_y_continuous(n.breaks = 4) +
    coord_cartesian(
        xlim = c(NA, 22),   # crop x axis limits (does not remove data points)
        clip = "off"        # prevent clipping of any labels
    ) +
  
    # set text for: title, subtitle, caption
    labs(
        title = str_c(
            "<b>Evening Taxi Trips in NYC are Dominated by ",
            "<span style='color:", COLOUR_UBER,
            "; font-weight:bold;'>Uber</span></b>"      # hardcoded vendor
        ),
        subtitle = "Hourly distribution of <b>taxi trips</b> by vendor, NYC (2021)",
        caption = FIG_CAPTION,
        x = "Pickup Hour", 
        y = NULL    
    ) +
    
    # apply base theme
    theme_minimal(
        base_family = BASE_FAMILY,
        base_size = BASE_SIZE   
    ) +
  
    # finetune theme elements
    theme(
        # title, subtitle, caption
        plot.title = element_markdown(          # markdown elements to add HTML/CSS styling/colouring
            size = rel(1.5),                    # relative font size to base_size (from base theme)
            margin = margin(b = 12)             # bottom vertical spacing
        ),
        plot.subtitle = element_markdown(
            colour = "#6e6e6e",               # grey
            size = rel(1),
            margin = margin(b = 24),
        ),
        plot.caption = element_text(
            colour = "#6e6e6e", 
            size = rel(0.7),
            hjust = 1                           # caption right-aligned (0 = left-aligned)
        ),
        plot.title.position = "plot",           # align title to the left of the figure
        # axes, grid, legend, margin
        axis.title.x = element_text(margin = margin(t = 9)),
        panel.grid.major.y = element_blank(),   
        panel.grid.minor = element_blank(),     # remove all minor grid lines
        axis.text.x = element_text(             # x axis ticker labels
            vjust = 0.5,   # distance to x axis
            colour = "black",
            size = rel(1.2)
        ),
        legend.position = "none",               # remove all legends
        plot.margin = margin(15, 15, 15, 15),   # set consistent outer border space
    )
```

```{r}
#| include: false

# save figure output to file
ggsave(
    file.path(FIGURES, "pickup_hour_activity_by_vendor.svg"),
    width  = knitr::opts_current$get("fig.width"),
    height = knitr::opts_current$get("fig.height")
)
```

:::